---
title: "Chapter3_Bacteria"
output:
  html_document:
    df_print: paged
  word_document: default
  pdf_document: default
editor_options:
  chunk_output_type: console
---
 <style type="text/css">

code.r{
  font-size: 12px;
}
pre {
  font-size: 10px
}
</style>

## 1. Loading packages
```{r message=FALSE}
library(ape)
library(vegan)
library(plyr)
library(dplyr)
library(scales)
library(grid)
library(reshape2)
library(phyloseq)
library(magrittr)
library(ggplot2)
library(ggpubr)
library(data.table)
library(tidyverse)
library(car)
library(BiocManager)
library(devtools)
library(lme4)
library(performance)
library(car)
library(emmeans)
library(plot3D)
library(viridis)
```

## 2. Loading the data
```{r}
otu <- read.table(file = "final_feature-table_filtered.tsv", sep = "\t", header = T, row.names = 1, 
                  skip = 1, comment.char = "")
taxonomy <- read.table(file = "taxonomy.tsv", sep = "\t", header = T ,row.names = 1)
taxonomy[,2]<-NULL #delete confidence column

metadata <- read.table(file = "sample_metadata_R.tsv", sep = "\t", header = T, row.names = 1)
metadata <- metadata %>%
  mutate(Treatment_label = case_when(
    Treatment == "G1" ~ "Early DSE",
    Treatment == "G2" ~ "Synchronous arrival",
    Treatment == "G3" ~ "No priority, young",
    Treatment == "G4" ~ "No priority, old",
    Treatment == "G5" ~ "Early DSE, no AMF",
    Treatment == "G6" ~ "No priority, no AMF, young",
    Treatment == "G7" ~ "No priority, no AMF, old",
    TRUE ~ as.character(Treatment)))

OTU = otu_table(as.matrix(otu), taxa_are_rows = TRUE)
TAX = tax_table(as.matrix(taxonomy))
SAMPLE <- sample_data(metadata)
TREE = read_tree("tree.nwk")

ps <- phyloseq(OTU, TAX, SAMPLE,TREE)
ps


#remove samples that are from the filter paper
samples_to_remove <- c("PEBAC221", "PEBAC222", "PEBAC223", "PEBAC224", "PEBAC225", "PEBAC226", "PEBAC227",
                       "PEBAC228", "PEBAC229", "PEBAC229", "PEBAC230", "PEBAC231", "PEBAC232", "PEBAC233",
                       "PEBAC234", "PEBAC235", "PEBAC236")
ps2 <- subset_samples(ps, !(SampleID %in% samples_to_remove))
ps2

#NEW object with ONLY the removed samples
ps_removed <- subset_samples(ps, SampleID %in% samples_to_remove)
ps_removed
```

## 3. Create rarefaction curve

```{r}
tab <- otu_table(ps2) 
class(tab) <- "matrix"
tab <- t(tab)
rarecurve(tab, step=20, lwd=2, ylab="OTU",  label=F)
```

## 4. Rarefy to minimum 70819, create rarefaction curve
```{r}
sample_sums <- sample_sums(ps2)
sums<-as.data.frame(sample_sums)
min(sample_sums) #70819
max((sample_sums)) #239936

total_reads <- sum(sample_sums(ps2))
#total_reads 29,912,057

set.seed(111) #keep results reproducible
ps.rarefied = rarefy_even_depth(ps2, rngseed=1, sample.size=70819, replace=F)
ps.rarefied
saveRDS(ps.rarefied, file = "phyloseq_rarefied.rds")
```

```{r}
tab2 <- otu_table(ps.rarefied)
class(tab2) <- "matrix"
tab2 <- t(tab2)
rarecurve(tab2, step=20, lwd=2, ylab="OTU",  label=F)
```

## 5. Alpha diversity analyses

 - shannon: How many kinds & how balanced are they? not very sensitive to dominant species.
 - simpson: How much do the common ones dominate?, more sensitive if there are taxa that dominates. 
 - so if a community is not well balanced, simpson will drop more than shannon. although the direction will be the same. but simpson will have a lower value compared to simpson. 
 
### 5.1. Estimate richness
```{r}
richness <- estimate_richness(ps.rarefied)
data_merge <- merge(richness, metadata, 
                    by = 'row.names', all = TRUE)
data_merge<-na.omit(data_merge)

data_merge <- data_merge %>%
  mutate(Pielou = Shannon / log(Observed),
         Pielou = ifelse(is.finite(Pielou), Pielou, NA_real_))
```

#### 5.2. Testing if early DSE arrival has an effect in diversity, same approach as in Biomass

- We will test this by comparing treatment G1 vs G3.
- To test if this effect is truly because of DSE and not AMF, we will compare G1 vs G5.
- This analysis answer the question: Does DSE early arrival has an effect in bacterial diversity?

##### Shannon diversity

```{r}
data_merge$Block<-as.factor(data_merge$Block) 
data_merge$Treatment<-as.factor(data_merge$Treatment)

Early<-filter(data_merge,Treatment %in% c("G1", "G3", "G5"))

#block explain ~ 0% variance
e1<-lmer(Shannon~Treatment + (1|Block), data = Early) 
check_model(e1)
summary(e1)
Anova(e1, type = 2)

emm <- emmeans(e1, ~ Treatment)
pairs(emm, adjust = "FDR") 
```

##### Simpson diversity

```{r}
#block explain ~ 0% variance
e2<- lmer(Simpson~Treatment + (1|Block), data=Early)
check_model(e2)
summary(e2)
Anova(e2, type = 2)

emm <- emmeans(e2, ~ Treatment)
pairs(emm, adjust = "FDR") 
```

##### Richness

```{r}
#block explain ~ 5.61% variance
e3<-lmer(Observed~Treatment + (1|Block), data=Early)
check_model(e3)
summary(e3)
Anova(e3, type = 2)

emm <- emmeans(e3, ~ Treatment)
pairs(emm, adjust = "FDR") 
```

##### Pielou

```{r}
#block explain ~ 0 variance
e4<-lmer(Pielou~Treatment + (1|Block), data=Early)
check_model(e4)
summary(e4)
Anova(e4, type = 2)

emm <- emmeans(e4, ~ Treatment)
pairs(emm, adjust = "FDR") 
```

```{r, fig.width=18, fig.height= 10}
p1<-ggerrorplot(Early, x="Treatment_label", y="Observed",
            xlab = "Treatment", 
            ylab= "Richness",
            size = 1) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 14)) +
  theme(
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 10)) 

p2<-ggerrorplot(Early, x="Treatment_label", y="Shannon",
            xlab = "Treatment", 
            ylab= "Shannon diversity index",
            size = 1) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 14)) +
  theme(
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 10)) 

p3<-ggerrorplot(Early, x="Treatment_label", y="Simpson",
            xlab = "Treatment", 
            ylab= "Simpson diversity index",
            size = 1) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 14)) +
  theme(
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 10))

p4<-ggerrorplot(Early, x="Treatment_label", y="Pielou",
            xlab = "Treatment", 
            ylab= "Pielou evenness index",
            size = 1) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 14)) +
  theme(
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 10))

g<-ggarrange(p1, p2, p3,p4, 
          ncol = 2, nrow = 2, 
          labels = c("A", "B", "C", "D")) 
annotate_figure(g,
  top = text_grob("Effect of early DSE arrival, compare G1-G3, G1-G5", face = "bold", size = 14))
```


```{r, fig.width=18, fig.height= 10}

p1<-ggerrorplot(data_merge, x="Treatment_label", y="Observed",
            xlab = "Treatment", 
            ylab= "Richness",
            size = 1) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 14)) +
  theme(
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 10)) 

p2<-ggerrorplot(data_merge, x="Treatment_label", y="Shannon",
            xlab = "Treatment", 
            ylab= "Shannon diversity index",
            size = 1) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 14)) +
  theme(
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 10)) 

p3<-ggerrorplot(data_merge, x="Treatment_label", y="Simpson",
            xlab = "Treatment", 
            ylab= "Simpson diversity index",
            size = 1) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 14)) +
  theme(
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 10))

p4<-ggerrorplot(data_merge, x="Treatment_label", y="Pielou",
            xlab = "Treatment", 
            ylab= "Pielou evenness index",
            size = 1) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 14)) +
  theme(
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 10))

g<-ggarrange(p1, p2, p3,p4, 
          ncol = 2, nrow = 2, 
          labels = c("A", "B", "C", "D")) 
annotate_figure(g,
  top = text_grob("Diversity metrics in Bacteria", face = "bold", size = 14))
```

- G1- G3: sig effect in Shannon/Simpson/Richness/Pielou. In fungi, sig. dif in Shannon/Simpson/Pielou but not in species richness. DSE arriving early decreases diversity and how evenly species are distributed (more uneven). So, it's possible DSE has an inhibitory effect/competitive in other microbes colonizing. Very similar than in fungi. 

- G1-G5: sig. dif. in  Shannon /Simpson/Pielou but not Richness. In fungi, sig dif in Shannon/Simpson/Pielou/richness. This means AMF decreases diversity by making the abundances of species more uneven, not how many different microbes are recruited. So AMF inhibits other microbes. Why? AMF could change root exudates. Competition for plant resources. 

- early DSE arrival affects bacterial richness but not fungal. 
- Having AMF affects fungal richness but not bacterial richness.

- the overall conclusion is the same in fungi and bacteria. 


#### 5.2.1 Does AMF/DSE identity matter in diversity if DSE arrives early?

- comparing G1 Vs G3 in four diversity measures

##### Shannon

```{r, fig.width=14, fig.height= 10}
Earlyd<-filter(data_merge,Treatment %in% c("G1", "G3"))
#Block explain ~1.39% of variance 
e6<-lmer(Shannon~Treatment*DSE_identity*AMF_identity+ (1|Block), data = Earlyd)
check_model(e6)
summary(e6)
Anova(e6, type = 2) # DSE identity, treatment and Treatment:AMF_identity

emm <- emmeans(e6, ~ DSE_identity)
pairs(emm, adjust = "FDR")

emm <- emmeans(e6, ~ DSE_identity|Treatment)
pairs(emm, adjust = "FDR")

emm <- emmeans(e6, ~ Treatment*AMF_identity)
pairs(emm, adjust = "FDR")

mycols <- c("DSE1" = "#007", 
             "DSE2" = "#D56")

mycols2 <- c("AMF1" = "#0072B2", 
             "AMF2" = "#D55E00")

a1 <-ggerrorplot(Earlyd, x="Treatment_label", y="Shannon", color = "DSE_identity",
            palette = mycols,
            xlab = "Treatment", 
            ylab= "Shannon diversity index",
            size = 1) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 14)) +
  theme(
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14)) 


a2 <-ggerrorplot(Earlyd, x="Treatment_label", y="Shannon", color = "AMF_identity",
            palette = mycols2,
            xlab = "Treatment", 
            ylab= "Shannon diversity index",
            size = 1) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 14)) +
  theme(
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14)) 

g<-ggarrange(a1, a2, 
          ncol = 2, nrow = 1, 
          labels = c("A", "B")) 
annotate_figure(g,
  top = text_grob("AMF/DSE identity in diversity if DSE arrives early", face = "bold", size = 14))

```

- DSE identity is important, treatment and Treatment:AMF_identity in Shannon diversity.In fungi, only treatment was significant
- DSE 2 is better than DSE1
- AMF1 in G1 is better than AMF2 in G1
- AMF1 in G3 is better than AMF2 in G1
- AMF2 in G3 is better than AMF2 in G1
- DSE2 in G3 is better than DSE2 in G1

##### Simpson

```{r, fig.width=14, fig.height= 10}
#Block explain ~0% of variance 
e7<-lmer(Simpson~Treatment*DSE_identity*AMF_identity+ (1|Block), data = Earlyd)
check_model(e7)
summary(e7)
Anova(e7, type = 2) # only treatment is significant and Treatment:AMF_identity

emm <- emmeans(e7, ~ Treatment*AMF_identity)
pairs(emm, adjust = "FDR")

mycols <- c("DSE1" = "#007", 
             "DSE2" = "#D56")

mycols2 <- c("AMF1" = "#0072B2", 
             "AMF2" = "#D55E00")

a3 <-ggerrorplot(Earlyd, x="Treatment_label", y="Simpson", color = "AMF_identity",
            palette = mycols2,
            xlab = "Treatment", 
            ylab= "Simpson diversity index",
            size = 1) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 14)) +
  theme(
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14))

annotate_figure(a3,
  top = text_grob("AMF/DSE identity in diversity if DSE arrives early", face = "bold", size = 14))

```

- Treatment and Treatment:AMF_identity is important in Simpson. In fungal only treatment was important. 
- AMF1 in G1 is better than AMF2 in G1.
- AMF1 in G3 is better than AMF2 in G1.
- AMF2 in G3 is better than AMF2 in G1

##### Pielou

```{r, fig.width=14, fig.height= 10}
#Block explain ~0% of variance 
e8<-lmer(Pielou~Treatment*DSE_identity*AMF_identity+ (1|Block), data = Earlyd)
check_model(e8)
summary(e8)
Anova(e8, type = 2) # treatment is significant, Treatment:AMF_identity

emm <- emmeans(e8, ~ Treatment*AMF_identity)
pairs(emm, adjust = "FDR")

mycols <- c("DSE1" = "#007", 
             "DSE2" = "#D56")

mycols2 <- c("AMF1" = "#0072B2", 
             "AMF2" = "#D55E00")

a4<-ggerrorplot(Earlyd, x="Treatment_label", y="Pielou", color = "AMF_identity",
            palette = mycols2,
            xlab = "Treatment", 
            ylab= "Pielou evenness index",
            size = 1) +
  theme(
    axis.text = element_text(size = 14),  
    axis.title = element_text(size = 14))
annotate_figure(a4,
  top = text_grob("AMF/DSE identity in diversity if DSE arrives early", face = "bold", size = 14))
```

- Treatment and Treatment:AMF_identity is important in Pielou (same as in Simpson). In fungal only treatment was important. 
- AMF1 in G1 is better than AMF2 in G1.
- AMF1 in G3 is better than AMF2 in G1.
- AMF2 in G3 is better than AMF2 in G1

##### Richness

```{r, fig.width=14, fig.height= 10}
#Block explain ~21.15% of variance 
e9<-lmer(Observed~Treatment*DSE_identity*AMF_identity+ (1|Block), data = Earlyd)
check_model(e9)
summary(e9)
Anova(e9, type = 2) # treatment is significant, DSE identity, Treatment:AMF_identity

emm <- emmeans(e9, ~ DSE_identity)
pairs(emm, adjust = "FDR")

emm <- emmeans(e9, ~ DSE_identity|Treatment)
pairs(emm, adjust = "FDR")

emm <- emmeans(e9, ~ Treatment*AMF_identity)
pairs(emm, adjust = "FDR")

mycols <- c("DSE1" = "#007", 
             "DSE2" = "#D56")

mycols2 <- c("AMF1" = "#0072B2", 
             "AMF2" = "#D55E00")

a5 <-ggerrorplot(Earlyd, x="Treatment_label", y="Observed", color = "DSE_identity",
            palette = mycols,
            xlab = "Treatment", 
            ylab= "Richness",
            size = 1) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 14)) +
  theme(
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14))


a6 <-ggerrorplot(Earlyd, x="Treatment_label", y="Observed", color = "AMF_identity",
            palette = mycols2,
            xlab = "Treatment", 
            ylab= "Richness",
            size = 1) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 14)) +
  theme(
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14))

g<-ggarrange(a5, a6, 
          ncol = 2, nrow = 1, 
          labels = c("A", "B")) 
annotate_figure(g,
  top = text_grob("AMF/DSE identity in diversity if DSE arrives early", face = "bold", size = 14))
```

- DSE identity is important, treatment and Treatment:AMF_identity in richness.In fungi, only treatment was signifficant
- DSE2 in G3 is better than DSE2 in G1.
- DSE2 in G3 is better than DSE1 in G3

- for AMF*treatment, all pairwise comparison are significant except AMF1 in G1 - AMF1 in G3

##### Does AMF identity matter when we add AMF? 

- we will test in Shannon/Simpson/Pielou. Not sure if this is necessary 

```{r}
#block does not explain any variance
sub_G1 <- subset(Early, Treatment == "G1")

#block explains 0%
m_G1 <- lmer(Shannon ~ AMF_identity + (1|Block), data = sub_G1)
check_model(m_G1)
summary(m_G1)
Anova(m_G1, type = 2)# AMF matters
emm_m_G1 <- emmeans(m_G1, ~ AMF_identity)
pairs(emm_m_G1, adjust = "FDR")

#block explain ~0 of variance 
m_G1_2 <- lmer(Simpson ~ AMF_identity + (1|Block), data = sub_G1)
check_model(m_G1_2)
summary(m_G1_2)
Anova(m_G1_2, type = 2) #AMF matters
emm_m_G1_2 <- emmeans(m_G1_2, ~ AMF_identity)
pairs(emm_m_G1_2, adjust = "FDR")

#block explain ~0 of variance 
m_G1_3 <- lmer(Pielou ~ AMF_identity + (1|Block), data = sub_G1)
check_model(m_G1_3)
summary(m_G1_3)
Anova(m_G1_3, type = 2) #AMF matters
emm_m_G1_3 <- emmeans(m_G1_3, ~ AMF_identity)
pairs(emm_m_G1_3, adjust = "FDR")
```


### 5.3 Testing if synchronous arrival has an effect.

- We will test this by comparing treatment G1 vs G2, G2 vs G3.
- To test if this effect is truly because of DSE and not AMF, we will compare G2 vs G5.

#### 5.3.1 Synchronous arrival effect in diversity

- This analysis answer the question: Does synchronous has an effect in bacterial diversity?

##### Richness

```{r}
Synchronous<-filter(data_merge,Treatment %in% c("G1","G2","G3", "G5"))

#block explains 14.18% of variance
s1<-lmer(Observed~Treatment + (1|Block), data = Synchronous)
check_model(s1)
summary(s1) 
Anova(s1, type = 2)

emm1 <- emmeans(s1, ~ Treatment)
pairs(emm1, adjust = "FDR") 

```

##### Shannon diversity

```{r}
#block explains 6.69% of variance
s2<-lmer(Shannon~Treatment + (1|Block), data = Synchronous)
check_model(s2)
summary(s2) 
Anova(s2, type = 2)

emm2 <- emmeans(s2, ~ Treatment)
pairs(emm2, adjust = "FDR") 
```

##### Simpson diversity

```{r}
#block explains 5.82% of variance
s3<-lmer(Simpson~Treatment + (1|Block), data = Synchronous)
check_model(s3)
summary(s3) 
Anova(s3, type = 2)

emm3 <- emmeans(s3, ~ Treatment)
pairs(emm3, adjust = "FDR") 
```

##### Pielou

```{r}
#block explains 3.38% of variance
s4<-lmer(Pielou~Treatment + (1|Block), data = Synchronous)
check_model(s4)
summary(s4) 
Anova(s4, type = 2)

emm4 <- emmeans(s4, ~ Treatment)
pairs(emm4, adjust = "FDR") 
```

```{r, fig.width=18, fig.height= 10}
p5<-ggerrorplot(Synchronous, x="Treatment_label", y="Observed",
            xlab = "Treatment", 
            ylab= "Richness",
            size = 1) +
  theme(
    axis.text = element_text(size = 10),    # Increases the size of axis numbers
    axis.title = element_text(size = 10)    # Increases the size of axis labels
  ) 

p6<-ggerrorplot(Synchronous, x="Treatment_label", y="Shannon",
            xlab = "Treatment", 
            ylab= "Shannon diversity index",
            size = 1) +
  theme(
    axis.text = element_text(size = 10),    # Increases the size of axis numbers
    axis.title = element_text(size = 10)    # Increases the size of axis labels
  ) 

p7<-ggerrorplot(Synchronous, x="Treatment_label", y="Simpson",
            xlab = "Treatment", 
            ylab= "Simpson diversity index",
            size = 1) +
  theme(
    axis.text = element_text(size = 10),    # Increases the size of axis numbers
    axis.title = element_text(size = 10)    # Increases the size of axis labels
  ) 

p8<-ggerrorplot(Synchronous, x="Treatment_label", y="Pielou",
            xlab = "Treatment", 
            ylab= "Pielou evenness index",
            size = 1) +
  theme(
    axis.text = element_text(size = 10),    # Increases the size of axis numbers
    axis.title = element_text(size = 10)    # Increases the size of axis labels
  ) 

g<-ggarrange(p5, p6, p7,p8, 
          ncol = 2, nrow = 2, 
          labels = c("A", "B", "C", "D")) 
annotate_figure(g,
  top = text_grob("Effect of Synchronous arrival, compare G1-G2, G2-G3, G2-G5", face = "bold", size = 14))
```

- G1-G2: Not sig. dif in richness/Shannon/Simpson/Pielou. In fungi Sig. difference in Shannon/Pielou but not Simpson and richness. So for bacterial communities, inoculating DSE/AMF together early does not have priority effects. It does not matter if we inoculate early together or separate. Even if they compete with each other (DSE/AMF) they don't affect bacterial niches. But G2 is a little less than G1. So AMF may have a small but not significant effect.

- G2-G3: Sig. dif. in richness/Shannon/Simpson/Pielou. In fungi No sig. diff in Richness/Shannon/Simpson/Pielou. All other microbes in the full community (meaning all other bacteria) preempt?  AMF and DSE if arrive 2 weeks ealy, resulting in priority effects? or the difference is because all other microbes in the community have 2 more weeks in the roots, so the diversity and abundance is higher (difference is because more time in host).

- G2-G5: Sig. dif. in Shannon/Simpson/Pielou but not in richness. In fungi Sig. dif in Richness/Shannon/Simpson but not Pielou. Having AMF decreases diversity by decreasing the abundance of species in the community, not the number of species (opposite in fungi). This confirms AMF preempt other species hability to establish in the community (same as un fungi)


### 5.3.2 Does AMF/DSE identity matter in diversity when DSE and AMF arrive together?

- Not paying attention to G1-G2 because the general model says no difference 
- focusing on G2-G3


##### Richness

```{r, fig.width=14, fig.height= 10}
Synchronousd<-filter(data_merge,Treatment %in% c("G2", "G3"))

#Block explain 15.2% variance
s5<-lmer(Observed~Treatment*DSE_identity*AMF_identity + (1|Block), data = Synchronousd)
check_model(s5)
summary(s5)
Anova(s5, type = 2) #treatment, Treatment:DSE_identity, Treatment:AMF_identity

emm5 <- emmeans(s5, ~ DSE_identity)
pairs(emm5, adjust = "FDR")

emm6 <- emmeans(s5, ~ Treatment*DSE_identity)
pairs(emm6, adjust = "FDR")

emm7 <- emmeans(s5, ~ Treatment*AMF_identity)
pairs(emm7, adjust = "FDR")

mycols <- c("DSE1" = "#007", 
             "DSE2" = "#D56")

mycols2 <- c("AMF1" = "#0072B2", 
             "AMF2" = "#D55E00")

a7 <-ggerrorplot(Synchronousdd, x="Treatment_label", y="Observed", color = "DSE_identity",
            palette = mycols,
            xlab = "Treatment", 
            ylab= "Richness",
            size = 1) +
  theme(
    axis.text = element_text(size = 14),  
    axis.title = element_text(size = 14))


a8 <-ggerrorplot(Synchronousdd, x="Treatment_label", y="Observed", color = "AMF_identity",
            palette = mycols2,
            xlab = "Treatment", 
            ylab= "Richness",
            size = 1) +
  theme(
    axis.text = element_text(size = 14),  
    axis.title = element_text(size = 14))

g<-ggarrange(a7, a8, 
          ncol = 2, nrow = 1, 
          labels = c("A", "B")) 
annotate_figure(g,
  top = text_grob("AMF/DSE identity in diversity if DSE/AMF arrive together", face = "bold", size = 14))

```

- For DSE identity
- DSE2 is better than DSE1

- For Treatment:DSE
- DSE2 in G3 is better than DSE1 in G2
- DSE2 in G3 is better than DSE2 in G2
- DSE2 in G3 is better than DSE1 in G3

- For Treatment:AMF
- AMF2 in G3 is better than AMF1 in G2
- AMF1 in G3 is better than AMF2 in G2
- AMF2 in G3 is better than AMF2 in G2
- AMF2 in G3 is better than AMF1 in G3

##### Shannon

```{r}
#Block explain 11.3% variance
s6<-lmer(Shannon~Treatment*DSE_identity*AMF_identity + (1|Block), data = Synchronousd)
check_model(s6)
summary(s6)
Anova(s6, type = 2) #DSE identity and Treatment:AMF_identity 

emm8 <- emmeans(s6, ~ DSE_identity)
pairs(emm8, adjust = "FDR")

emm9 <- emmeans(s6, ~ Treatment*AMF_identity)
pairs(emm9, adjust = "FDR")

mycols <- c("DSE1" = "#007", 
             "DSE2" = "#D56")

mycols2 <- c("AMF1" = "#0072B2", 
             "AMF2" = "#D55E00")

a9 <-ggerrorplot(Synchronousdd, x="Treatment_label", y="Shannon", color = "DSE_identity",
            palette = mycols,
            xlab = "Treatment", 
            ylab= "Shannon diversity index",
            size = 1) +
  theme(
    axis.text = element_text(size = 14),  
    axis.title = element_text(size = 14))


a10 <-ggerrorplot(Synchronousdd, x="Treatment_label", y="Shannon", color = "AMF_identity",
            palette = mycols2,
            xlab = "Treatment", 
            ylab= "Shannon diversity index",
            size = 1) +
  theme(
    axis.text = element_text(size = 14),  
    axis.title = element_text(size = 14))

g<-ggarrange(a9, a10, 
          ncol = 2, nrow = 1, 
          labels = c("A", "B")) 
annotate_figure(g,
  top = text_grob("AMF/DSE identity in diversity if DSE/AMF arrive together", face = "bold", size = 14))
```

- For DSE identity
- DSE2 is better than DSE1

- For Treatment:AMF
- AMF1 in G3 is better than AMF1 in G2
- AMF2 in G3 is better than AMF1 in G2
- AMF1 in G3 is better than AMF2 in G2
- AMF2 in G3 is better than AMF2 in G2

##### Simpson

```{r, fig.width=14, fig.height= 10}
#Block explain 8.8% variance
s7<-lmer(Simpson~Treatment*DSE_identity*AMF_identity + (1|Block), data = Synchronousd)
check_model(s7)
summary(s7)
Anova(s7, type = 2) #treatment, Treatment:AMF_identity, DSE_identity:AMF_identity

emm10 <- emmeans(s7, ~ Treatment*AMF_identity)
pairs(emm10, adjust = "FDR") 

emm11 <- emmeans(s7, ~ DSE_identity*AMF_identity) #nothing here
pairs(emm11, adjust = "FDR") 

emm12 <- emmeans(s7, ~ AMF_identity | DSE_identity)
pairs(emm12, adjust="FDR") #this includes all three treatments, in the plot only G2 and G3

emm13 <- emmeans(s7, ~ DSE_identity | AMF_identity) #this makes more sense according to plots?
pairs(emm13, adjust="FDR") #this includes all three treatments, in the plot only G2 and G3

mycols <- c("DSE1" = "#007", 
             "DSE2" = "#D56")

mycols2 <- c("AMF1" = "#0072B2", 
             "AMF2" = "#D55E00")

p<-ggerrorplot(Synchronousdd, x="Treatment_label", y="Simpson", color = "AMF_identity",
            palette = mycols2,
            xlab = "Treatment", 
            ylab= "Simpson diversity index",
            size = 1)+
  labs(title = "DSE/AMF identity if DSE/AMF arrive together") +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.text = element_text(size = 14),  
    axis.title = element_text(size = 14))
p
p + facet_wrap(~DSE_identity)

pp<-ggerrorplot(Synchronousdd, x="Treatment_label", y="Simpson", color = "DSE_identity",
            palette = mycols,
            xlab = "Treatment", 
            ylab= "Simpson diversity index",
            size = 1)+
  labs(title = "DSE/AMF identity if DSE/AMF arrive together") +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.text = element_text(size = 14),  
    axis.title = element_text(size = 14))

pp + facet_wrap(~AMF_identity)

```

- For Treatment:AMF
- AMF1 in G3 is better than AMF1 in G2
- AMF2 in G3 is better than AMF1 in G2
- AMF1 in G3 is better than AMF2 in G2
- AMF2 in G3 is better than AMF2 in G2

- For DSE_identity*AMF_identity, no specific pairwise comparison

- For AMF_identity | DSE_identity
-within DSE1, overall AMF1 is better than AMF2 (not looking at treatment)
-nothing within DSE2

- For DSE_identity | AMF_identity
- nothing within AMF1
- within AMF2, DSE2 is better than DSE1

##### Pielou

```{r, fig.width=14, fig.height= 10}
#Block explain 4.7% variance
s8<-lmer(Pielou~Treatment*DSE_identity*AMF_identity + (1|Block), data = Synchronousd)
check_model(s8)
summary(s8)
Anova(s8, type = 2) #treatment, Treatment:AMF_identity

emm13 <- emmeans(s8, ~ Treatment*AMF_identity)
pairs(emm13, adjust = "FDR") 

mycols2 <- c("AMF1" = "#0072B2", 
             "AMF2" = "#D55E00")

p2 <- ggerrorplot(Synchronousdd, x="Treatment_label", y="Pielou", color="AMF_identity",
                  palette = mycols2,
                  xlab = "Treatment",
                  ylab = "Pielou evenness index",
                  size = 1) +
  labs(title = "DSE/AMF identity if DSE/AMF arrive together") +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.text  = element_text(size = 14),
    axis.title = element_text(size = 14))

p2
```

- For Treatment:AMF_identity
- AMF1 in G3 is better than AMF1 in G2
- AMF2 in G3 is better than AMF1 in G2
- AMF1 in G3 is better than AMF2 in G2
- AMF2 in G3 is better than AMF2 in G2


##### Does AMF identity matter when we add AMF? 

- we will test in Shannon/Simpson/Pielou. Not sure if this is necessary 

```{r}
#block does not explain any variance
Synchronousdd<-filter(data_merge,Treatment %in% c("G2", "G5"))
sub_G2 <- subset(Synchronousd, Treatment == "G2")

#block explains 28.3%
m_G2 <- lmer(Shannon ~ AMF_identity + (1|Block), data = sub_G2)
check_model(m_G2)
summary(m_G2)
Anova(m_G2, type = 2)# not important

#block explain ~9.9 of variance 
m_G2_2 <- lmer(Simpson ~ AMF_identity + (1|Block), data = sub_G2)
check_model(m_G2_2)
summary(m_G2_2)
Anova(m_G2_2, type = 2) #AMF does not matter

#block explain ~15.1 of variance 
m_G2_3 <- lmer(Pielou ~ AMF_identity + (1|Block), data = sub_G2)
check_model(m_G2_3)
summary(m_G2_3)
Anova(m_G2_3, type = 2) #AMF does not matter

```

- it does not matter the identity of AMF. We just know that it inhibits colonization of other microbes (same as in fungi). 

### 5.4 Testing if host age has an effect and not arrival order.

- We will test this by comparing treatment G3 vs G4 (including AMF), G6 vs G7 (only the DSE)
- To test if this effect is truly because of plant age and not the AMF, we will compare G3 vs G6, G4 vs G7

#### 5.4.1 Host age in bacterial diversity

- This analysis answer the question: Does host age has an effect in fungal diversity?

##### Richness

```{r}
Hostage<-filter(data_merge,Treatment %in% c("G3","G4","G6", "G7"))

#block explains ~8.09%
hs1<-lmer(Observed~Treatment + (1|Block), data = Hostage)
check_model(hs1)
summary(hs1) 
Anova(hs1, type = 2)

emm14 <- emmeans(hs1, ~ Treatment)
pairs(emm14, adjust = "FDR") 
```

##### Shannon

```{r}
#block explains ~5.88%
hs2<-lmer(Shannon~Treatment + (1|Block), data = Hostage)
check_model(hs2)
summary(hs2) 
Anova(hs2, type = 2)

emm15 <- emmeans(hs2, ~ Treatment)
pairs(emm15, adjust = "FDR") 
```

##### Simpson

```{r}
#block explains ~0%
hs3<-lmer(Simpson~Treatment + (1|Block), data = Hostage)
check_model(hs3)
summary(hs3) 
Anova(hs3, type = 2)

emm16 <- emmeans(hs3, ~ Treatment)
pairs(emm16, adjust = "FDR")

```

##### Pielou

```{r}
#block explains ~0.78%
hs4<-lmer(Pielou~Treatment + (1|Block), data = Hostage)
check_model(hs4)
summary(hs4) 
Anova(hs4, type = 2)

emm <- emmeans(hs4, ~ Treatment) # is significant but not the combination we want
pairs(emm, adjust = "FDR") 
```


```{r, fig.width=18, fig.height= 10}
p9<-ggerrorplot(Hostage, x="Treatment_label", y="Observed",
            xlab = "Treatment", 
            ylab= "Richness",
            size = 1) +
  theme(
    axis.text = element_text(size = 10),    
    axis.title = element_text(size = 10)) 

p10<-ggerrorplot(Hostage, x="Treatment_label", y="Shannon",
            xlab = "Treatment", 
            ylab= "Shannon diversity index",
            size = 1) +
  theme(
    axis.text = element_text(size = 10),    
    axis.title = element_text(size = 10)) 

p11<-ggerrorplot(Hostage, x="Treatment_label", y="Simpson",
            xlab = "Treatment", 
            ylab= "Simpson diversity index",
            size = 1) +
  theme(
    axis.text = element_text(size = 10),  
    axis.title = element_text(size = 10)) 

p12<-ggerrorplot(Hostage, x="Treatment_label", y="Pielou",
            xlab = "Treatment", 
            ylab= "Pielou evenness index",
            size = 1) +
  theme(
    axis.text = element_text(size = 10),   
    axis.title = element_text(size = 10)) 

g<-ggarrange(p9, p10, p11,p12, 
          ncol = 2, nrow = 2, 
          labels = c("A", "B", "C", "D")) 
annotate_figure(g,
  top = text_grob("Effect of Host age, compare G3-G4, G6-G7, G3-G7, G7-G4", face = "bold", size = 14))
```

- G3-G4: not sig.dif in richness/Shannon/Simpson/Pielou. In fungi sig. dif. in Shannon but not in Richness/Simpson/Pielou: Host age has no effect in the relative abundance of species or richness. There is not dominant taxa. all relatively the same. This treatment includes DSE/AMF/full community. 

- G6-G7: Sig. dif. in richness not in Shannon/Simpson/Pielou. In fungi, sig. dif in Richness/Shannon/Simpson/Pielou: when there is not AMF, host age affects only richness. Because more time for more species to colonize roots? But overall, no influence of age in species abundance and evennnes. 

- G3-G6: Sig. dif in richness/Shannon not in Simpson/Pielou. In fungi, sig. dif in Richness/Shannon/Simpson/Pielou. G6 does not have AMF. Having AMF confirms priority effects by AMF, inhibitory (same in fungi) 

- G7-G4: Sig. dif in richness not in Shannon/Simpson/Pielou. In fungi, sig. dif in Richness/Shannon not in Simpson/Pielou. G7 does not have AMF.  

#### 5.4.2 Does AMF/DSE identity matter bacterial diversity?

- Not doing this, because models says no difference. 

## 6. PERMANOVAS

```{r}
set.seed(20)
bray <- phyloseq::distance(ps.rarefied, method="bray") #bray distance matrix of abundance
adonis2(bray ~ sample_data(ps.rarefied)$Treatment, by = "terms", permutations = 1000)

bray_mat <- as.matrix(bray) 
bray_df <- as.data.frame(as.matrix(bray))

library(pairwiseAdonis)

treat <- factor(metadata$Treatment)

#pairwise.adonis2(bray_df, factors = treat, p.adjust.m = "BH")

library(RVAideMemoire)

fact <- factor(sample_data(ps.rarefied)$Treatment)
pairwise.perm.manova(bray, fact = fact, nperm = 1000, p.method = "BH")


mod1 <- betadisper(bray, phyloseq::sample_data(ps.rarefied)$Treatment)
mod1
permutest(mod1, permutations =999)

pair.test.mod1<-permutest(mod1, pairwise = TRUE, permutations = 99) 
pair.test.mod1

## Extract distances + group from betadisper object
df_mod1 <- data.frame(
  Day  = mod1$group,
  Dist = mod1$distances)

df_mod1 <- df_mod1 %>%
  mutate(Treatment_label = case_when(
    Day == "G1" ~ "Early DSE",
    Day == "G2" ~ "Synchronous arrival",
    Day == "G3" ~ "No priority, young",
    Day == "G4" ~ "No priority, old",
    Day == "G5" ~ "Early DSE, no AMF",
    Day == "G6" ~ "No priority, no AMF, young",
    Day == "G7" ~ "No priority, no AMF, old",
    TRUE ~ as.character(Day)))


ggerrorplot(df_mod1, x="Treatment_label", y="Dist",
            xlab = "Assembly history", 
            ylab= "Distance to group centroid",
            size = 1) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 14)) +
  theme(
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14))

```

```{r}
nmds.bray <- ordinate(ps.rarefied, method = "NMDS", distance = "bray")
nmds.bray
stressplot(nmds.bray)

plot_ordination(ps.rarefied, nmds.bray, color = "Treatment")

nmds1<-plot_ordination(ps.rarefied, nmds.bray, color = "Treatment", shape = "AMF_identity")
nmds1
nmds1 + facet_wrap(~DSE_identity)+theme_bw()
nmds1 + facet_wrap(~AMF_identity)+theme_bw()

nmds2<-plot_ordination(ps.rarefied, nmds.bray, color = "Treatment", shape = "DSE_identity")
nmds2
nmds2 + facet_wrap(~AMF_identity)+theme_bw()
nmds2 + facet_wrap(~DSE_identity)+theme_bw()
```

## 7. Relative abundance plots

- First instead of having the names out of dada2, lets change the rownames to OTU_001....
- Then we will separate the taxonomy table in columns

```{r}
ps_rel<-ps.rarefied
ps_rel
tax_table(ps_rel) <- cbind(tax_table(ps_rel),#copies rownames, this variable is called automatically V8
                                           rownames(tax_table(ps_rel)))

tax_ps_rel<-as.data.frame(tax_table(ps_rel)) #check if worked
colnames(tax_table(ps_rel)) <-c("Taxonomy", "ASV") #to change V2 to OTU
taxa_names(ps_rel) <- sprintf("%04d", 1:ntaxa(ps_rel)) #we change row names in taxonomy table to ASV_0001, ASV_0002...
taxa_names(ps_rel) <- paste("ASV_",taxa_names(ps_rel),sep = "")
tax_ps_rel<-as.data.frame(tax_table(ps_rel)) #extract tax table to see if worked
tax_ps_rel2<-subset(tax_ps_rel, select = Taxonomy)

extract_family <- function(tax_string) {
  taxa <- unlist(strsplit(tax_string, ";"))  # Split taxonomy string
  
  # Trim whitespace and extract taxonomic ranks
  taxa <- trimws(taxa)  
  family <- taxa[grepl("^f__", taxa)]
  order <- taxa[grepl("^o__", taxa)]
  class <- taxa[grepl("^c__", taxa)]
  phylum <- taxa[grepl("^p__", taxa)]

  # Check if family is available and valid
  if (length(family) > 0 && family != "f__" && family != "f__ ") {
    return(sub("f__", "", family))  # Remove prefix
  }
  
  # Fallback to order, class, or phylum if family is missing
  if (length(order) > 0 && order != "o__" && order != "o__ ") {
    return(sub("o__", "", order))
  } else if (length(class) > 0 && class != "c__" && class != "c__ ") {
    return(sub("c__", "", class))
  } else if (length(phylum) > 0 && phylum != "p__" && phylum != "p__ ") {
    return(sub("p__", "", phylum))
  } else {
    return("unidentified")  # Default if no classification is found
  }
}

# Apply function to extract new names based on family
new_names <- sapply(tax_ps_rel2$Taxonomy, extract_family)

# Rename row names to include family instead of species
rownames(tax_ps_rel2) <- paste0(rownames(tax_ps_rel2), "_", new_names)

# Extract the OTU table and update row names
otu_table <- as.data.frame(otu_table(ps_rel))  # Extract the OTU table

otu_old_names <- rownames(otu_table)  # Original OTU names
otu_new_names <- rownames(tax_ps_rel2)  # Updated OTU names with family

# Ensure both have the same OTU order before renaming
if (length(otu_old_names) == length(otu_new_names)) {
  names(otu_new_names) <- otu_old_names  # Create a named vector mapping
  rownames(otu_table) <- otu_new_names[rownames(otu_table)]  # Apply new names
} else {
  stop("Error: The number of OTUs in otu_table and tax_ps_guilds2 does not match.")
}

# Verify renaming worked
identical(rownames(otu_table), rownames(tax_ps_rel2))  # Should return TRUE


metadata<-as.data.frame(sample_data(ps_rel)) #extract metadata from ps object

OTU = otu_table(as.matrix(otu_table), taxa_are_rows = TRUE)
TAX = tax_table(as.matrix(tax_ps_rel2))
SAMPLE <- sample_data(metadata)

Bacteria<- phyloseq(OTU, TAX, SAMPLE)
Bacteria

```

- get taxonomy separated

```{r}
#create a copy of Bacteria
Bacteria2<-Bacteria

#extract tax table
tax_table_traits<-as.data.frame(tax_table(Bacteria2))

#Separate in columns
tax <- tax_table_traits %>%
  select(Taxonomy) %>%
  separate(Taxonomy, into = c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species"), sep = ";")

tax.clean <- data.frame(
  row.names = row.names(tax),
  Kingdom = str_replace(tax[,1], "d__", ""),
  Phylum = str_replace(tax[,2], "p__", ""),
  Class = str_replace(tax[,3], "c__", ""),
  Order = str_replace(tax[,4], "o__", ""),
  Family = str_replace(tax[,5], "f__", ""),
  Genus = str_replace(tax[,6], "g__", ""),
  Species = str_replace(tax[,7], "s__", ""),
  stringsAsFactors = FALSE
)

tax.clean[is.na(tax.clean)] <- ""
tax.clean[tax.clean == "__"] <- ""

for (i in 1:nrow(tax.clean)) {
  if (tax.clean[i, 2] == "") {
    kingdom <- paste("Unclassified", tax.clean[i, 1], sep = " ")
    tax.clean[i, 2:7] <- kingdom
  } else if (tax.clean[i, 3] == "") {
    phylum <- paste("Unclassified", tax.clean[i, 2], sep = " ")
    tax.clean[i, 3:7] <- phylum
  } else if (tax.clean[i, 4] == "") {
    class <- paste("Unclassified", tax.clean[i, 3], sep = " ")
    tax.clean[i, 4:7] <- class
  } else if (tax.clean[i, 5] == "") {
    order <- paste("Unclassified", tax.clean[i, 4], sep = " ")
    tax.clean[i, 5:7] <- order
  } else if (tax.clean[i, 6] == "") {
    family <- paste("Unclassified", tax.clean[i, 5], sep = " ")
    tax.clean[i, 6:7] <- family
  } else if (tax.clean[i, 7] == "") {
    # If species is missing but genus exists, fill Species with "Unclassified <Genus>"
    tax.clean$Species[i] <- paste("Unclassified", tax.clean[i, 6], sep = " ")
  }
}

tax.clean2 <- as.matrix(tax.clean) #convert to matrix
tax.clean3 <- tax_table(tax.clean2) #convert to tax table for phyloseq
tax_table(Bacteria2) <- tax.clean3
#saveRDS(Bacteria2, file = "Bacteria_ASV00_final.rds") #this has the otus changed to ASV_0001
```

#### 7.2 Relative abundance plots

```{r}
#Bac_barplot <- readRDS("Bacteria_ASV00_final.rds")
#ps_family <- tax_glom(Bac_barplot, taxrank = "Family") #ps object of only genus
#ps_family

#data_family_type <- psmelt(ps_family) #data frame simplified to genus of abundances
#data_family_type <- as.data.frame(psmelt(ps_family))

#write_csv(data_family_type, "data_family_type.csv") #calling this is much easier than doing all the above

data_family_type2<-read.csv("data_family_type.csv")

# per-sample relative abundance
smp_ra <- data_family_type2 %>%
  group_by(Sample, UniqueTreatment, OTU) %>%
  summarise(Count = sum(Abundance, na.rm = TRUE), .groups = "drop_last") %>%
  mutate(RelAbund = 100 * Count / sum(Count)) %>%
  ungroup()

# average within UniqueTreatment
grp_ra <- smp_ra %>%
  group_by(UniqueTreatment, OTU) %>%
  summarise(mean_RA = mean(RelAbund, na.rm = TRUE),
            se_RA   = sd(RelAbund, na.rm = TRUE)/sqrt(dplyr::n()),
            n       = dplyr::n(),
            .groups = "drop")

# collapse low genera (<4%) AFTER averaging
thresh <- 4
grp_ra <- grp_ra %>%
  mutate(OTU = ifelse(mean_RA < thresh, "Other", as.character(OTU))) %>%
  group_by(UniqueTreatment, OTU) %>%
  summarise(mean_RA = sum(mean_RA), .groups = "drop") %>%
  mutate(OTU = fct_relevel(OTU, "Other", after = Inf))

# x-axis order (use order present in data)
x_levels <- grp_ra %>%
  dplyr::distinct(UniqueTreatment) %>%
  dplyr::mutate(
    G = as.integer(sub("^G(\\d+)\\..*$", "\\1", UniqueTreatment)),
    R = as.integer(sub("^G\\d+\\.(\\d+)$", "\\1", UniqueTreatment))
  ) %>%
  dplyr::arrange(G, R) %>%
  dplyr::pull(UniqueTreatment)

grp_ra$UniqueTreatment <- factor(grp_ra$UniqueTreatment, levels = x_levels)

# stacked barplot
library(pals)
my_14_colors <- alphabet(26)

p <- ggplot(grp_ra, aes(x = UniqueTreatment, y = mean_RA, fill = OTU)) +
  geom_col(width = 0.8) +
  labs(x = NULL, y = "Relative abundance (%)", fill = "Genus/OTU") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 0.4))

p + scale_fill_manual(values = unname(alphabet(26)))

```

#### 7.3 The barplot of the inoculum

- it may be a good idea to join both phyloseq objects once both have been rarefied, but we need to delete the tree.

```{r}
sample_sums <- sample_sums(ps_removed)
sums<-as.data.frame(sample_sums)
min(sample_sums) #133488, but will rarefied to same number of root samples
max((sample_sums)) #187660

total_reads <- sum(sample_sums(ps_removed))
#total_reads 2,621,975

set.seed(111) #keep results reproducible
ps.rarefied2 = rarefy_even_depth(ps_removed, rngseed=1, sample.size=70819, replace=F)
ps.rarefied2

ps_rarefied_notree <- phyloseq(otu_table(ps.rarefied), tax_table(ps.rarefied), sample_data(ps.rarefied))
ps_rarefied_notree

ps_rarefied2_notree <- phyloseq(otu_table(ps.rarefied2), tax_table(ps.rarefied2), sample_data(ps.rarefied2))
ps_rarefied2_notree

ps_rare_both <- merge_phyloseq(ps_rarefied_notree, ps_rarefied2_notree)
ps_rare_both

```

- Change dada2 names to ASV_001 of both ps objects

```{r}
ps_rel<-ps_rare_both
ps_rel

tax_table(ps_rel) <- cbind(tax_table(ps_rel),#copies rownames, this variable is called automatically V8
                                           rownames(tax_table(ps_rel)))

tax_ps_rel<-as.data.frame(tax_table(ps_rel)) #check if worked
colnames(tax_table(ps_rel)) <-c("Taxonomy", "ASV") #to change V2 to OTU
taxa_names(ps_rel) <- sprintf("%04d", 1:ntaxa(ps_rel)) #we change row names in taxonomy table to ASV_0001, ASV_0002...
taxa_names(ps_rel) <- paste("ASV_",taxa_names(ps_rel),sep = "")
tax_ps_rel<-as.data.frame(tax_table(ps_rel)) #extract tax table to see if worked
tax_ps_rel2<-subset(tax_ps_rel, select = Taxonomy)

extract_family <- function(tax_string) {
  taxa <- unlist(strsplit(tax_string, ";"))  # Split taxonomy string
  
  # Trim whitespace and extract taxonomic ranks
  taxa <- trimws(taxa)  
  family <- taxa[grepl("^f__", taxa)]
  order <- taxa[grepl("^o__", taxa)]
  class <- taxa[grepl("^c__", taxa)]
  phylum <- taxa[grepl("^p__", taxa)]

  # Check if family is available and valid
  if (length(family) > 0 && family != "f__" && family != "f__ ") {
    return(sub("f__", "", family))  # Remove prefix
  }
  
  # Fallback to order, class, or phylum if family is missing
  if (length(order) > 0 && order != "o__" && order != "o__ ") {
    return(sub("o__", "", order))
  } else if (length(class) > 0 && class != "c__" && class != "c__ ") {
    return(sub("c__", "", class))
  } else if (length(phylum) > 0 && phylum != "p__" && phylum != "p__ ") {
    return(sub("p__", "", phylum))
  } else {
    return("unidentified")  # Default if no classification is found
  }
}

# Apply function to extract new names based on family
new_names <- sapply(tax_ps_rel2$Taxonomy, extract_family)

# Rename row names to include family instead of species
rownames(tax_ps_rel2) <- paste0(rownames(tax_ps_rel2), "_", new_names)

# Extract the OTU table and update row names
otu_table <- as.data.frame(otu_table(ps_rel))  # Extract the OTU table

otu_old_names <- rownames(otu_table)  # Original OTU names
otu_new_names <- rownames(tax_ps_rel2)  # Updated OTU names with family

# Ensure both have the same OTU order before renaming
if (length(otu_old_names) == length(otu_new_names)) {
  names(otu_new_names) <- otu_old_names  # Create a named vector mapping
  rownames(otu_table) <- otu_new_names[rownames(otu_table)]  # Apply new names
} else {
  stop("Error: The number of OTUs in otu_table and tax_ps_guilds2 does not match.")
}

# Verify renaming worked
identical(rownames(otu_table), rownames(tax_ps_rel2))  # Should return TRUE


metadata<-as.data.frame(sample_data(ps_rel)) #extract metadata from ps object

OTU = otu_table(as.matrix(otu_table), taxa_are_rows = TRUE)
TAX = tax_table(as.matrix(tax_ps_rel2))
SAMPLE <- sample_data(metadata)

Bacteria_both<- phyloseq(OTU, TAX, SAMPLE)
Bacteria_both

```

- get taxonomy separated in columns of both ps objects

```{r}
#create a copy of Bacteria_both
Bacteria2<-Bacteria_both

#extract tax table
tax_table_traits<-as.data.frame(tax_table(Bacteria2))

#Separate in columns
tax <- tax_table_traits %>%
  select(Taxonomy) %>%
  separate(Taxonomy, into = c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species"), sep = ";")

tax.clean <- data.frame(
  row.names = row.names(tax),
  Kingdom = str_replace(tax[,1], "d__", ""),
  Phylum = str_replace(tax[,2], "p__", ""),
  Class = str_replace(tax[,3], "c__", ""),
  Order = str_replace(tax[,4], "o__", ""),
  Family = str_replace(tax[,5], "f__", ""),
  Genus = str_replace(tax[,6], "g__", ""),
  Species = str_replace(tax[,7], "s__", ""),
  stringsAsFactors = FALSE
)

tax.clean[is.na(tax.clean)] <- ""
tax.clean[tax.clean == "__"] <- ""

for (i in 1:nrow(tax.clean)) {
  if (tax.clean[i, 2] == "") {
    kingdom <- paste("Unclassified", tax.clean[i, 1], sep = " ")
    tax.clean[i, 2:7] <- kingdom
  } else if (tax.clean[i, 3] == "") {
    phylum <- paste("Unclassified", tax.clean[i, 2], sep = " ")
    tax.clean[i, 3:7] <- phylum
  } else if (tax.clean[i, 4] == "") {
    class <- paste("Unclassified", tax.clean[i, 3], sep = " ")
    tax.clean[i, 4:7] <- class
  } else if (tax.clean[i, 5] == "") {
    order <- paste("Unclassified", tax.clean[i, 4], sep = " ")
    tax.clean[i, 5:7] <- order
  } else if (tax.clean[i, 6] == "") {
    family <- paste("Unclassified", tax.clean[i, 5], sep = " ")
    tax.clean[i, 6:7] <- family
  } else if (tax.clean[i, 7] == "") {
    # If species is missing but genus exists, fill Species with "Unclassified <Genus>"
    tax.clean$Species[i] <- paste("Unclassified", tax.clean[i, 6], sep = " ")
  }
}

tax.clean2 <- as.matrix(tax.clean) #convert to matrix
tax.clean3 <- tax_table(tax.clean2) #convert to tax table for phyloseq
tax_table(Bacteria2) <- tax.clean3
#saveRDS(Bacteria2, file = "Bacteria_ASV00_final_both.rds") #this has the otus changed to ASV_0001
```

- Plots, family level

```{r, fig.width=20, fig.height= 10}
#ASV_barplot <- readRDS("Bacteria_ASV00_final_both.rds")
#ps_family <- tax_glom(ASV_barplot, taxrank = "Family") #ps object of only genus
#ps_family

#data_family_type <- psmelt(ps_family) #data frame simplified to genus of abundances
#data_family_type <- as.data.frame(psmelt(ps_family))

#write_csv(data_family_type, "data_family_type_both.csv") #calling this is much easier than doing all the above

data_family_type2<-read.csv("data_family_type_both.csv")

# per-sample relative abundance by Family
smp_ra <- data_family_type2 %>%
  group_by(Sample, UniqueTreatment, Family) %>%
  summarise(Count = sum(Abundance, na.rm = TRUE), .groups = "drop_last") %>%
  mutate(RelAbund = 100 * Count / sum(Count)) %>%
  ungroup()

# average within UniqueTreatment
grp_ra <- smp_ra %>%
  group_by(UniqueTreatment, Family) %>%
  summarise(mean_RA = mean(RelAbund, na.rm = TRUE),
            se_RA   = sd(RelAbund, na.rm = TRUE)/sqrt(dplyr::n()),
            n       = dplyr::n(),
            .groups = "drop")

# collapse low genera (<3%) AFTER averaging
thresh <- 3
grp_ra <- grp_ra %>%
  mutate(Family_collapsed = ifelse(mean_RA < thresh, "Other", as.character(Family))) %>%
  group_by(UniqueTreatment, Family = Family_collapsed) %>%
  summarise(mean_RA = sum(mean_RA), .groups = "drop") %>%
  mutate(Family = fct_relevel(Family, "Other", after = Inf))

# x-axis order (use order present in data)
x_levels <- grp_ra %>%
  dplyr::distinct(UniqueTreatment) %>%
  dplyr::mutate(
    G = as.integer(sub("^G(\\d+)\\..*$", "\\1", UniqueTreatment)),
    R = as.integer(sub("^G\\d+\\.(\\d+)$", "\\1", UniqueTreatment))
  ) %>%
  dplyr::arrange(G, R) %>%
  dplyr::pull(UniqueTreatment)

grp_ra$UniqueTreatment <- factor(grp_ra$UniqueTreatment, levels = x_levels)

# stacked barplot
library(pals)
Family_levels <- levels(grp_ra$Family)
pal <- setNames(pals::glasbey(length(Family_levels)), Family_levels)

p <- ggplot(grp_ra, aes(x = UniqueTreatment, y = mean_RA, fill = Family)) +
  geom_col(width = 0.8) +
  labs(x = NULL, y = "Mean Relative Abundance (%)", fill = "Family/other < 3%") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 0.4))

p + scale_fill_manual(values = pal, breaks = names(pal))
```

```{r, fig.width=20, fig.height= 10}
#pick the last two levels on the x-axis
last2 <- c("G8.1","G9.1") 

#add a panel label
grp_ra_panels <- grp_ra %>%
  mutate(Panel = ifelse(UniqueTreatment %in% last2,
                        "Inoculum", "Roots"))

#colorblind-friendly palette for all OTUs in the data
pal <- pals::glasbey(nlevels(grp_ra_panels$Family))
names(pal) <- levels(grp_ra_panels$Family)

# 4) plot with two panels; each panel shows only its x values
p <- ggplot(grp_ra_panels, aes(x = UniqueTreatment, y = mean_RA, fill = Family)) +
  geom_col(width = 0.8) +
  facet_grid(~ Panel, scales = "free_x", space = "free_x") +
  scale_fill_manual(values = pal) +
  labs(x = NULL, y = "Mean Relative Abundance (%)", fill = "Family/other <3%") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 0.4),
        strip.placement = "outside",
        panel.spacing = unit(1.2, "lines"))

p
```

- Plots, genus level


```{r, fig.width=20, fig.height= 10}
#ASV_barplot <- readRDS("Bacteria_ASV00_final_both.rds")
#ps_genus <- tax_glom(ASV_barplot, taxrank = "Genus") #ps object of only genus
#ps_genus

#data_genus_type <- psmelt(ps_genus) #data frame simplified to genus of abundances
#data_genus_type <- as.data.frame(psmelt(ps_genus))

#write_csv(data_genus_type, "data_genus_type_both.csv") #calling this is much easier than doing all the above

data_genus_type2<-read.csv("data_genus_type_both.csv")

# per-sample relative abundance
smp_ra <- data_genus_type2 %>%
  group_by(Sample, UniqueTreatment, Genus) %>%
  summarise(Count = sum(Abundance, na.rm = TRUE), .groups = "drop_last") %>%
  mutate(RelAbund = 100 * Count / sum(Count)) %>%
  ungroup()

# average within UniqueTreatment
grp_ra <- smp_ra %>%
  group_by(UniqueTreatment, Genus) %>%
  summarise(mean_RA = mean(RelAbund, na.rm = TRUE),
            se_RA   = sd(RelAbund, na.rm = TRUE)/sqrt(dplyr::n()),
            n       = dplyr::n(),
            .groups = "drop")

# collapse low genera (<3%) AFTER averaging
thresh <- 3
grp_ra <- grp_ra %>%
  mutate(Genus_collapsed = ifelse(mean_RA < thresh, "Other", as.character(Genus))) %>%
  group_by(UniqueTreatment, Genus = Genus_collapsed) %>%
  summarise(mean_RA = sum(mean_RA), .groups = "drop") %>%
  mutate(Genus = fct_relevel(Genus, "Other", after = Inf))

# x-axis order (use order present in data)
x_levels <- grp_ra %>%
  dplyr::distinct(UniqueTreatment) %>%
  dplyr::mutate(
    G = as.integer(sub("^G(\\d+)\\..*$", "\\1", UniqueTreatment)),
    R = as.integer(sub("^G\\d+\\.(\\d+)$", "\\1", UniqueTreatment))
  ) %>%
  dplyr::arrange(G, R) %>%
  dplyr::pull(UniqueTreatment)

grp_ra$UniqueTreatment <- factor(grp_ra$UniqueTreatment, levels = x_levels)

# stacked barplot
library(pals)
gen_levels <- levels(grp_ra$Genus)
pal <- setNames(pals::glasbey(length(gen_levels)), gen_levels)

p <- ggplot(grp_ra, aes(x = UniqueTreatment, y = mean_RA, fill = Genus)) +
  geom_col(width = 0.8) +
  labs(x = NULL, y = "Mean Relative Abundance (%)", fill = "Genus/other <3%") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 0.4))

p + scale_fill_manual(values = pal, breaks = names(pal))
```

- Separate the inoculum

```{r, fig.width=20, fig.height= 10}
last2 <- c("G8.1","G9.1")

grp_ra_panels <- grp_ra %>%
  mutate(Panel = ifelse(as.character(UniqueTreatment) %in% last2, "Inoculum", "Roots"))

gen_levels <- levels(grp_ra_panels$Genus)
pal <- setNames(pals::glasbey(length(gen_levels)), gen_levels)

p_panels <- ggplot(grp_ra_panels, aes(UniqueTreatment, mean_RA, fill = Genus)) +
  geom_col(width = 0.8) +
  facet_grid(~ Panel, scales = "free_x", space = "free_x") +
  # drop=TRUE so 'Inoculum' shows only G8.1 & G9.1
  scale_x_discrete(drop = TRUE) +
  scale_fill_manual(values = pal, breaks = gen_levels) +
  labs(x = NULL, y = "Mean Relative Abundance (%)", fill = "Genus / Other <3%") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 0.4),
        strip.placement = "outside",
        panel.spacing = unit(1.2, "lines"))
p_panels
```


